Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway

### 2021-07-26 analysis/paper/paper.Rmd
Introducing logistic/residential a bit more throughly

```
The Early Mesolithic, or Flake Axe Phase, is defined as lasting from c. 9200--8200 BCE [@reitan2016]. The phase is set to start with the first human occupation in Norway which is widely held as originating from southern Scandinavian and northern European regions. This is to be directly reflected by similarities in the artefact inventories [@bjerck2008; @bang-andersen2012; @fuglestvedt2012; @glørstad2016]. Previous research has typically proposed that the Early Mesolithic is characterised by a relatively high degree of mobility, and low variation in site types and associated mobility patterns [e.g. @bjerck2008; @breivik2016; @fuglestvedt2012; but see @viken2018]. The transition to the subsequent Middle Mesolithic, or Microlith Phase, at around 8200 BCE has been linked to changes in blade [@damlien2016] and subsequently axe technology [@eymundsson2018; @solheim2020], which in turn has been associated with changes in population genomics and related migration events hailing from the Eurasian steppes [@günther2018; @manninen2021]. The radiocarbon record points towards a coinciding population decline in southern Norway around this time [@nielsen2021]. Although this does not appear to be evident in the regional data for south-eastern Norway, taphonomic loss associated with these early dates is an issue [@solheim2018a; @solheim2020a; @nielsen2021]. In the chronological framework of @reitan2016, the Microlith phase is defined as lasting until around 7000 BCE. The Microlith phase is followed by the Pecked Adze Phase, characterised by a more dominating presence of non-flint macro tools and associated production waste in the assemblages [@reitan2016]. Following @reitan2016, the next typological transition at c. 5600 BCE signifies the onset of the Nøstvet Adze Phase. While previously defined as having a slightly longer duration, the Nøstvet Phase has traditionally been seen as representing the onset of more varied settlement systems and stable mobility patterns in the Mesolithic of south-eastern Norway [e.g. @jaksland2001; @lindblom1984] In recent years it has been suggested that the transition to a decrease in mobility and more varied land-use patterns can be traced back to the Middle Mesolithic [@solheim2016]. The subsequent Transverse Arrowhead Phase (c. 4500--3900 BCE) is characterised by a dramatic decrease in axe finds, and the introduction of tranverse-, tanged- and single-edged points [@reitan2016]. It has recently been suggested that a dispersal of people from southern Scandinavia into southern Norway takes place in this period [@eigeland2015, 379; @nielsen2021], which could follow after a preceding population decline at c. 4300 BCE [@nielsen2021].

@@

sites %>% 
  mutate(strt_bce = reported_start_bc * -1, 
         end_bce = reported_end_bc * -1,
         dating_method = ifelse(dating_method == "shore_typo", "Shoreline/typology", "Radiocarbon")) %>% 
  select(no, site, dating_method, strt_bce, end_bce) %>% 
  gt() %>% 
    cols_label(
    site = "Site name",
    dating_method = "Dating method",
    strt_bce = "Reported start (BCE)",
    end_bce = "Reported end (BCE)",
  ) %>% 
  tab_header(
    title = "Table 1. Analysed sites."
  ) %>% 
  tab_options(heading.title.font.size = 15)
```
# The analysis of lithic assemblages

@@

Furthermore, one concrete operationalisation of the terms have been forwarded by @barton1998 and collegues [e.g. @barton2013; @barton1999; @riel-salvatore2004; @riel-salvatore2007; @barton2011; @barton2014; @clark2017; @villaverde1998], who through a series of studies have shown that the relationship between volumetric density of lithics and relative frequency of retouched artefacts in lithic assemblages have a consistent negative relationship across a wide range of chronological and cultural context, ranging from Pleistocene and Holocene assemblages in Europe and Asia, to assemblages associated with both Neanderthals and modern humans [@barton2011; @riel-salvatore2008]. This relationship is taken to reflect degree of curation, and is in turn mainly to follow from the accumulated nature of land-use and mobility patterns associated with the assemblages [@barton2014]. In this model, higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliability, repairability, and the degree to which artefacts could be manipulated to fulfil a wide range of tasks are therefore assumed to have been factors of concern. From this it follows that the empirical expectation for short-term camps is a curated technological organisation with higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017]. More time spent in a single location, on the other hand, is assumed to lead to better control of raw-material availability and to allow for its accumulation. This should in turn lead to a more expedient technological organisation with reduced necessity for the conservation of lithics and extensive use of retouch. The empirical expectation for lower degree of mobility is therefore relatively high density of lithics, a low relative frequency of retouched artefacts, as well as a higher number of cores and unretouched flakes and blades. These variables and underlying logic constitute what has been termed the whole assemblage behavioural index [WABI, @clark2017], and is the main framework adopted here.

However, as these measures are argued to predominantly be determined by land-use and mobility patterns, relative frequency of chips and relative frequency of non-flint material are also included in the analysis, as these measures have also been linked to mobility patterns and is of central importance in Norwegian Stone Age archaeology [e.g. @Bicho2020; @breivik2016a; @reitan2016; @kitchel2021] --- the use of local non-flint material has been taken to indicate reduced mobility and increased familiarity with local surroundings [@jaksland2001, 112; @glørstad2010, 181].

@@

The relatively constrained geographical distribution of the analysed sites, the limited temporal range over which they where investigated, as well as the methodological equivalency across excavation projects hopefully leads to an exclusion of some biases that might otherwise skew an exploratory analysis, rendering it more likely that behaviourally meaningful patterns are identified. However, the exploratory perspective means that a wide range of combinations and transformations of variables has been explored to identify patterning in the data. While only parts of this process can sensibly be reported here, the data and employed R programming script is freely available as a research compendium on GitHub (https://github.com/isakro/exploring-assemblages-se-norway) and Zenodo ([placeholder]), following @marwick2018b, allowing readers to explore and scrutinise the data and the final analytical choices made [@Marwick2017]. However, this inductive data-dredging or pattern-searching approach does constitute a limited inferential framework [@clark2009], as it involves a *post hoc* accommodation of explanations to meet the observed data---data that is both selectively and subjectively reported upon. The process can still provide the identification of empirical patterns with respects to the employed units of analysis, which in turn can form the basis for social and behavioural hypotheses. This can lay the foundation for a deductive research agenda with targeted model evaluation for which clear test implications can be derived [@clark2009, 29].

@@

The column plot (Figure \@ref(fig:cor)B) reveals that the earliest sites are characterised by the flint artefact categories microburins, projectiles, as well as flint macro tools and associated debitage. It is also interesting that these sites to a larger extent are characterised by core fragments, both in flint and non-flint materials, rather than the cores themselves. The non-flint material on the earliest, or among the earliest sites, appears to be centred around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites.

@@

Site number 9, Nedre Hobekk 2, located in the upper right quadrant of the row plot represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite [@eigeland2014]. However, as the site had been quite heavily impacted by modern disturbances, this led @eigeland2014 [124] to suggest that the material might have been compromised. This could explain its position as an outlier in the plot. This could indicate that specialised axe production sites disappear towards the end of the Mesolithic, a notion that would be in line with previous research [e.g. @eigeland2015, 370; @glørstad2011; @reitan2016]. Apart from the impact of core fragments, which is not always highlighted, this must be considered very much in agreement with previous research [e.g. @solheim2017c, with references]. A marked presence of core fragments has, however, previously been noted as one of several similarities between Early Mesolithic Norwegian sites and Late Palaeolithic sites from continental Europe [@fuglestvedt2007; see also @bjerck2017]. Overall, the comparatively limited impact of the flint material can possibly be the result of the aggregation of categories that leads to an suppression of otherwise temporally distinct patterns, and there are certainly technological nuances in the flint material that is temporally contingent but not recorded during a regular classification of the material [e.g. @damlien2016; @eigeland2015; @fuglestvedt2007; @solheim2020]. However, while the former pertains to the analytical trade-off between robustness and sensitivity, the latter is likely to be true for the non-flint material as well [@eigeland2007]. The overall pattern does speak to the impact the properties of the raw-material has for the general composition of the assemblages [cf. @manninen2014].
The use of metarhyolite for the production of axes is present at other contemporary sites as well [@jaksland2014b], but is evidently not as prominent a part of these assemblages. In sum, the findings for the earliest sites are in large part in line with previous research [e.g @bjerck2017; @breivik2018; @fuglestvedt2007; @jaksland2014b; @damlien2018b].

The results of the CA does appear to align well with previous research, and the employed artefact categories are clearly capturing a temporal component. One possible implication of this close correspondence could be that the aggregation of artefact categories might have been overly conservative. However, it is also evidently clear, in the words of @kruskal1971 [22], that 'time is not the only dimension'. The results of the CA do most certainly correspond to more pervasive cultural change than a purely typo-chronological development of artefact morphology, which is also made evident by some of significant deviances from the overall pattern. Unpicking and aligning these patterns with any specific behavioural and technological dimensions using the coarse CA results is, however, another task entirely. This follows most clearly from the fact that for the most part we do not know what individual lithic objects in the assemblages has been used for, leaving the behavioural and social significance of the employed units of analysis unclear. The results of the CA can, however, be used in conjunction with the part of the analysis that has attempted to get at more specific behavioural dimensions to nuance or explain discrepancies in this data.

@@

The curation index then remains stable for the rest of the Mesolithic. This suggests that the transition to mobility patterns traditionally ascribed to the Nøstvet Phase can indeed be traced back to the Microlith Phase [@solheim2016]. The continued stability of the curation index could also indicate that the demographic changes suggested to take place in the Transverse Arrowhead Phase are not related to major shifts in land-use and mobility patterns in the material treated here. However, it is worth highlighting the strained sample size for the later parts of the Mesolithic, which could mean that the effect is simply missed, especially if the signal is weaker than that for the transition from the Early Mesolithic.

>>>>

The Early Mesolithic, or Flake Axe Phase, is defined as lasting from c. 9200--8200 BCE [@reitan2016]. The phase is set to start with the first human occupation in Norway which is widely held as originating from southern Scandinavian and northern European regions. This is to be directly reflected by similarities in the artefact inventories [@bjerck2008; @bang-andersen2012; @berg-hansen2018; @fuglestvedt2012; @glørstad2016]. Previous research has typically proposed that the Early Mesolithic is characterised by a relatively high degree of mobility, and low variation in site types and associated mobility patterns [e.g. @bjerck2008; @breivik2016; @fuglestvedt2012; but see @viken2018]. The transition to the subsequent Middle Mesolithic, or Microlith Phase, at around 8200 BCE has been linked to changes in blade [@damlien2016] and subsequently axe technology [@eymundsson2018; @solheim2020], which in turn has been associated with changes in population genomics and related migration events hailing from the Eurasian steppes [@günther2018; @manninen2021]. The radiocarbon record points towards a coinciding population decline in southern Norway around this time [@nielsen2021]. Although this does not appear to be evident in the regional data for south-eastern Norway, taphonomic loss associated with these early dates is an issue [@solheim2018a; @solheim2020a; @nielsen2021]. In the chronological framework of @reitan2016, the Microlith phase is defined as lasting until around 7000 BCE. The Microlith phase is followed by the Pecked Adze Phase, characterised by a more dominating presence of non-flint macro tools and associated production waste in the assemblages [@reitan2016]. Following @reitan2016, the next typological transition at c. 5600 BCE signifies the onset of the Nøstvet Adze Phase. While previously defined as having a slightly longer duration, the Nøstvet Phase has traditionally been seen as representing the onset of more varied settlement systems and stable mobility patterns in the Mesolithic of south-eastern Norway [e.g. @jaksland2001; @lindblom1984] In recent years it has been suggested that the transition to a decrease in mobility and more varied land-use patterns can be traced back to the Middle Mesolithic [@solheim2016]. The subsequent Transverse Arrowhead Phase (c. 4500--3900 BCE) is characterised by a dramatic decrease in axe finds, and the introduction of tranverse-, tanged- and single-edged points [@reitan2016]. It has recently been suggested that a dispersal of people from southern Scandinavia into southern Norway takes place in this period [@eigeland2015, 379; @nielsen2021], which could follow after a preceding population decline at c. 4300 BCE [@nielsen2021].

@@

sites %>% 
  mutate(strt_bce = reported_start_bc * -1, 
         end_bce = reported_end_bc * -1,
         dating_method = ifelse(dating_method == "shore_typo", "Shoreline/typology", "Radiocarbon")) %>% 
  select(no, site, dating_method, strt_bce, end_bce) %>% 
  gt() %>% 
    cols_label(
    site = "Site name",
    dating_method = "Dating method",
    strt_bce = "Reported start (BCE)",
    end_bce = "Reported end (BCE)",
  ) %>% 
  tab_header(
    title = "Table 1. Analysed sites."
  ) %>% 
  tab_options(heading.title.font.size = 15)
```

# The analysis of lithic assemblages

@@

Furthermore, one concrete operationalisation of the terms have been forwarded by @barton1998 and collegues [e.g. @barton2013; @barton1999; @riel-salvatore2004; @riel-salvatore2007; @barton2011; @barton2014; @clark2017; @villaverde1998], who through a series of studies have shown that the relationship between volumetric density of lithics and relative frequency of retouched artefacts in lithic assemblages have a consistent negative relationship across a wide range of chronological and cultural context, ranging from Pleistocene and Holocene assemblages in Europe and Asia, to assemblages associated with both Neanderthals and modern humans [@barton2011; @riel-salvatore2008]. This relationship is taken to reflect degree of curation, and is in turn mainly to follow from the accumulated nature of land-use and mobility patterns associated with the assemblages [@barton2014].

In this model, higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliability, repairability, and the degree to which artefacts could be manipulated to fulfil a wide range of tasks are therefore assumed to have been factors of concern. From this it follows that the empirical expectation for short-term camps is a curated technological organisation with higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017]. More time spent in a single location, on the other hand, is assumed to lead to better control of raw-material availability and to allow for its accumulation. This should in turn lead to a more expedient technological organisation with reduced necessity for the conservation of lithics and extensive use of retouch. The empirical expectation for lower degree of mobility is therefore relatively high density of lithics, a low relative frequency of retouched artefacts, as well as a higher number of cores and unretouched flakes and blades. These variables and underlying logic constitute what has been termed the whole assemblage behavioural index [WABI, @clark2017], and is the main framework adopted here.

However, as these measures are argued to predominantly be determined by land-use and mobility patterns, relative frequency of chips and relative frequency of non-flint material are also included in the analysis, as these measures have also been linked to mobility patterns and is of central importance in Norwegian Stone Age archaeology [e.g. @Bicho2020; @breivik2016a; @reitan2016; @kitchel2021]---the use of local non-flint material has been taken to indicate reduced mobility and increased familiarity with local surroundings [@jaksland2001, 112; @glørstad2010, 181].

This relationship has been related to the continuum defined by @binford1980 between residentially mobile foragers and logistically mobile collectors. Residential mobility involves the movement of entire groups between resource patches throughout the year, while logistic mobility entails the use of central base-camps from where task groups venture to retrieve resources, which can lead to a wider spectrum of site types and associated mobility patterns. While this framwork can offer

@@

The relatively constrained geographical distribution of the analysed sites, the limited temporal range over which they where investigated, as well as the methodological equivalency across excavation projects hopefully leads to an exclusion of some biases that might otherwise skew an exploratory analysis, rendering it more likely that behaviourally meaningful patterns are identified. However, the exploratory perspective means that a wide range of combinations and transformations of variables has been explored to identify patterning in the data. While only parts of this process can sensibly be reported here, the data and employed R programming script is freely available as a research compendium on GitHub (<https://github.com/isakro/exploring-assemblages-se-norway>) and Zenodo ([placeholder]), following @marwick2018b, allowing readers to explore and scrutinise the data and the final analytical choices made [@Marwick2017]. However, this inductive data-dredging or pattern-searching approach does constitute a limited inferential framework [@clark2009], as it involves a *post hoc* accommodation of explanations to meet the observed data---data that is both selectively and subjectively reported upon. The process can still provide the identification of empirical patterns with respects to the employed units of analysis, which in turn can form the basis for social and behavioural hypotheses. This can lay the foundation for a deductive research agenda with targeted model evaluation for which clear test implications can be derived [@clark2009, 29].

@@

The column plot (Figure \@ref(fig:cor)B) reveals that the earliest sites are characterised by the flint artefact categories microburins, projectiles, as well as flint macro tools and associated debitage. It is also interesting that these sites to a larger extent are characterised by core fragments, both in flint and non-flint materials, rather than the cores themselves. The non-flint material on the earliest, or among the earliest sites, appears to be centred around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites. Site number 9, Nedre Hobekk 2, located in the upper right quadrant of the row plot represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite [@eigeland2014]. However, as the site had been quite heavily impacted by modern disturbances, this led @eigeland2014 [124] to suggest that the material might have been compromised. This could explain its position as an outlier in the plot.

@@

The results of the CA does appear to align well with previous research [e.g. @solheim2017c, with references]. In the flint material the earliest sites are separated from the rest primarily based on the presence of macro tools, microburins, projectiles, and, for slightly younger sites, core fragments and blades [@bjerck2017; @breivik2018; @fuglestvedt2007; @eymundsson2018; @jaksland2014b; @damlien2018b]. The importance of the latter two can be associated with the blade technology that is introduced with the Middle Mesolithic, characterised by blade production from conical and sub-conical cores with faceted platforms that involves the removal of core tablets and rejuvenation flakes [e.g. @damlien2016]. When it comes to the non-flint material, projectiles are to a larger extent a property of the earlier sites than later ones. The use of metarhyolite for the production of axes is present at some earlier sites in addition to the previously mentioned Nedre Hobekk 2, and the production of hatches and core axes is introduced in the Microlith Phase [@eymundsson2018; @jaksland2014b; @reitan2016]. However, in agreement with the literature, this is evidently not as prominent a part of these assemblages [@reitan2016; @damlien2018b].

The flint material of the later sites are to larger extent characterised by micro-blades, which corresponds to the transition to micro-blade production from handle cores [e.g. @glørstad2010, 161--163; @solheim2020]. A more fragmented flint material, as indicated by the relative importance of flint chips and fragments, is also a previously noted property of some later Mesolithic, as well as early Neolithic sites [e.g. @fossum2017; @stokke2018a]. The most defining material for the later sites, however, is non-flint macro tools and associated debitage, which is dominating some of these assemblages. It was noted above that despite the low number of the absolute earliest sites, these appear to not be impacted by this material. This could indicate that specialised axe production sites disappear towards the end of the Mesolithic, a notion that would be in line with previous suggestions [e.g. @eigeland2015, 370; @glørstad2011; @reitan2016].

Overall, the comparatively limited impact of the flint material can possibly be the result of the aggregation of categories that leads to an suppression of otherwise temporally distinct patterns, and there are certainly technological nuances in the flint material that is temporally contingent but not recorded during a regular classification of the material [e.g. @damlien2016; @eigeland2015; @fuglestvedt2007; @solheim2020]. However, while the former pertains to the analytical trade-off between robustness and sensitivity, the latter is likely to be true for the non-flint material as well [@eigeland2007]. The overall pattern does nonetheless speak to the impact the properties of the raw-material has for the general composition of the assemblages [cf. @manninen2014].

One possible implication of the fact that the employed artefact categories are so clearly capturing a temporal component could consequently be that the aggregation of artefact categories might have been overly conservative. However, it is also evidently clear, in the words of @kruskal1971 [22], that 'time is not the only dimension'. The results of the CA do most certainly correspond to more pervasive cultural change than a purely typo-chronological development of artefact morphology, which is also made evident by some of significant deviances from the overall pattern. Unpicking and aligning these patterns with any specific behavioural and technological dimensions using the coarse CA results is, however, another task entirely. This follows most clearly from the fact that for the most part we do not know what individual lithic objects in the assemblages has been used for, leaving the behavioural and social significance of the employed units of analysis unclear. The results of the CA can, however, be used in conjunction with the part of the analysis that has attempted to get at more specific behavioural dimensions to nuance or explain discrepancies in this data.

@@

The curation index then remains stable for the rest of the Mesolithic. This suggests that the transition to mobility patterns traditionally ascribed to the Nøstvet Phase can indeed be traced back to the Microlith Phase [cf. @solheim2016]. The continued stability of the curation index could also indicate that the demographic changes suggested to take place in the Transverse Arrowhead Phase are not related to major shifts in land-use and mobility patterns in the material treated here. However, it is worth highlighting the strained sample size for the later parts of the Mesolithic, which could mean that the effect is simply missed, especially if the signal is weaker than that for the transition from the Early Mesolithic.
